Healthcare Insights and Sociodemographic Analysis of New York: A Comprehensive Analysis of Patient Records in 2013

### 2023-10-08 NewSTAT405.qmd
Update NewSTAT405.qmd

*(last updated 09/30/2023)*

@@

```
<span style="font-size: 14px; font-style: italic;">Figure 1a: Length of Stay on Price Regression Dotplot</span>

@@

cat("\\begin{center}\n")
print(table_lmfs, comment = FALSE)
cat("\\end{center}\n")
```
<i>Figure 1b: Table of Linear Regression Modeling Data</i>

@@

```
<i>Figure 2: Total Costs Boxplot</i>

@@

```{r, warning = FALSE, fig.height = 6, fig.width = 50}
#| echo: false
# Set the margin to include more space at the bottom
par(mar = c(5, 4, 4, 2) + 0.1)

cleaned_data$`Payment Typology 1` <- as.factor(cleaned_data$`Payment Typology 1`)
payments <- table(cleaned_data$`Payment Typology 1`)
pie(payments, main = "Payment Type Distribution", col = rainbow(length(payments)), labels = NA)
legend(1, 1, 1, legend = levels(cleaned_data$`Payment Typology 1`), fill = rainbow(length(payments)), cex = 0.7)

```
<i>Figure 3: Pie Chart of Payment Distribution</i>

@@

grid()
```
<span style="font-size: 14px; font-style: italic;">Figure 4:Trends of Total Charges and Legnth of Stay </span>

@@

```
<span style="font-size: 14px; font-style: italic;">Figure 5: Age Group Distribution of Patients in New York</span>

@@

## **Plot 6: Heatmap of Gender vs. Age Group**

```{r}
#| echo: false
heatmap_data <- table(cleaned_data$`Age Group`, cleaned_data$Gender)

row_names <- as.factor(rownames(heatmap_data))
col_names <- as.factor(colnames(heatmap_data))

heatmap(heatmap_data,
    Rowv = NA,
    Colv = NA,
    col = colorRampPalette(c("white", "blue"))(50),
    xlab = "Gender",
    main = "Gender v. Age Group",
    labRow = row_names,
    labCol = col_names
)

```
<span style="font-size: 14px; font-style: italic;">Figure 6: Gender vs Age Group Heatmapt</span>

The following heatmap shows the relationship between the gender and age group for patients in New York in 2013. As we can observe, women tend to be patients at hospitals at a higher rate then men. Especially for the 70 or older bracket which is the highest frequency of hospital patients, females are significantly a larger proprotion of who are patients compared to men. Another important data point to understand is the identification of "U" or individuals who do not identify as male and female, which is high for ages 18-29, which is an important data point to help increase healthcare for individuals with fluid gender identity or may identify as non-binary.

## Plot 7: Stacked Bar Chart by Gender and Race

@@

plot7 <- ggplot(cleaned_data, aes(x = Gender, fill = Race)) +
  geom_bar() +
  labs(
    title = "Stacked Bar Chart by Gender and Race",
    x = "Gender",
    y = "Count"
  ) +
  scale_fill_manual(
    values = c("White" = "lightblue", "African American" = "lightcoral", "Other" = "lightgreen"),
    breaks = c("White", "Black", "Other"),
    labels = c("White", "Black", "Other"),
    name = "Race"
  ) +
  theme_minimal()

plot7 + ggtitle("Data Source: mtcars dataset, ggplot2 package")

```
<span style="font-size: 14px; font-style: italic;">Figure 7: Stacked Bar Chart by Race and Gender</span>

@@

## Plot 8: Boxplots on the Total Costs on the Risk of Mortality

@@

```
<span style="font-size: 14px; font-style: italic;">Figure 8: Boxplots on Total Cost and Mortality</span>

## Plot Description and Inferencing
In Figure 8, our objective was to assess the central tendency of total hospital costs across different levels of Risk of Mortality. Employing separate box plots for each severity category allowed us to discern distinct patterns in the distribution of total costs associated with varying degrees of mortality risk. Notably, when focusing on Extreme cases of mortality, a clear trend emerged—median total costs were significantly higher, accompanied by a broader spread in the distribution. This observation implies that patients facing an Extreme risk of mortality not only incur higher median costs but also exhibit greater variability in the financial implications of their healthcare, suggesting a more diverse range of medical interventions and resource utilization.

This finding aligns with our initial expectation that higher levels of risk of mortality correspond to elevated costs of care. The discernible trend in the box plots reinforces the intuitive understanding that patients with a greater severity of illness often require more extensive and resource-intensive medical interventions, leading to increased healthcare costs. Beyond its immediate implications, this nuanced analysis offers valuable insights for healthcare planning and resource allocation, emphasizing the need for tailored strategies based on the specific needs associated with varying risk levels. This approach contributes to more effective and efficient healthcare delivery, promoting a patient-centered and economically sustainable healthcare system.

@@

## Plot 9: Violin Plot Comparing New York City and Western New York's Severity of Illness

@@

```
<span style="font-size: 14px; font-style: italic;">Figure 9: NYC and Western NY Severity of Illness Comparsion </span>

## Plot Description and Inferencing
Figure 9 delves into a detailed breakdown of hospital patients, distinguishing between those originating from New York City (NYC) and Western New York (WNY). The focus is on the distribution of patients' APR severity codes, measured on a 1 to 4 scale (1 being mild, 2 being moderate, 3 being serious, and 4 being extreme). Notably, the violin plot reveals a higher frequency of APR severity code 2 assignments for patients from NYC compared to WNY. This divergence prompts exploration into potential variations in labeling practices between the two regions. However, beyond this distinction, the overall shapes of the violin distributions appear similar, suggesting a general alignment in severity profiles. This observation implies a degree of consistency in patient conditions based on APR severity codes, despite the regional differences in labeling practices. Understanding such variations is crucial for refining medical coding precision and ensuring uniformity in patient severity assessments across diverse geographical contexts.

@@

## Plot 10: Distribution of APR Severity of Illness for Medicaid Patients

@@

```
<span style="font-size: 14px; font-style: italic;">Figure 10: APR Severity of Illness for Medicaid Patients </span>

## Plot Description and Inferencing
In our examination of Figure 10, our focus shifted toward a specific subset of patients—those under Medicaid care. Medicaid, a health program designed to support individuals classified as economically disadvantaged, prompted our investigation into the severity of diagnoses within this demographic. Our objective was to discern whether patients under Medicaid receive comparable attention and care in contrast to their counterparts outside of the program.

@@

## Plot 11: Facet Grid of Dotplots for Total Charges on Length of Stay

```{r warning = FALSE}
#| echo: false
library(ggplot2)

ggplot(cleaned_data, aes(x = cleaned_data$`Length of Stay`, y = cleaned_data$`Total Charges`)) +
  geom_point(size = 0.2) +
  facet_grid(Race ~ `Type of Admission`) +
  labs(title = "Length of Stay v. Total Charges", x = "Length of Stay", y = "Total Charges") +
  scale_y_continuous(labels = scales::comma, 
                     breaks = scales::pretty_breaks(), 
                     limits = c(0, max(cleaned_data$`Total Charges`)), # 
                     expand = c(0, 0)) +
  theme(axis.text = element_text(size = 8))

```
<span style="font-size: 14px; font-style: italic;">Figure 11: Dotplots for Total Charges on Length of Stay </span>

@@

In Figure 11, our aim was to dissect assessments based on both race and the type of admission. Through our analysis, a discernible pattern emerged, revealing consistent spikes in each category of total charges, particularly towards the lower end of the length of stay spectrum. This observation suggests that, irrespective of race, there are commonalities in the distribution of total charges, with pronounced concentration points occurring for shorter stays.

Upon closer examination, the data points appear to be uniformly distributed across all racial categories, implying a degree of similarity in the cost distribution patterns among different racial groups. Interestingly, the primary differentiating factor contributing to variations in total costs is identified as the type of admission. This finding underscores the significance of admission characteristics in influencing healthcare costs, emphasizing that differences in total charges are more prominently linked to variations in admission types rather than racial distinctions.

These insights shed light on the nuanced interplay between race, type of admission, and healthcare costs. Understanding these relationships is crucial for healthcare administrators and policymakers in designing targeted strategies for cost-effective care delivery, ensuring equitable access, and fostering a healthcare system that is both inclusive and economically sustainable.

@@

## Plot 12: Facet Density Plots of Total Cost per Age-Group Buckets

gg + scale_y_continuous(labels = scales::number_format(scale = 1e-3)) +
  labs(title = "Density Plot of Total Costs per Age-Group Bucket", x = "Total Costs", y = "Density (in thousands)") +
  theme_minimal()
```
<span style="font-size: 14px; font-style: italic;">Figure 12: Density Plots of Total Costs per Age-Groups</span>

## Plot Description and Inferencing
In Figure 12, we analyzed age-specific total costs by charting density plots using a facet grid. For patients under 17, costs skewed right, suggesting lower expenses, potentially due to the overall good health of children. In contrast, as age increases, a broader and more dispersed peak in total costs emerges, indicating that older patients require more extensive and potentially costlier healthcare interventions, such as hospice services. This exploration sheds light on age-related patterns in healthcare costs, guiding strategies for resource allocation and care planning across different life stages.

### References
New York State Department of Health. "Hospital Inpatient Discharges (SPARCS De-Identified)." data.ny.gov, 2013, https://health.data.ny.gov/Health/Hospital-Inpatient-Discharges-SPARCS-De-Identified/npsr-cm47.

>>>>

*(last updated 10/07/2023)*

@@

```
<p align = "center">
[Figure 1a: Length of Stay on Price Regression Dotplot]
</p>

@@

cat("\\begin{center}\n")
print(table_lmfs, comment = FALSE)
cat("\\end{center}\n")
```
<p align = "center">
[Figure 1b: Table of Linear Regression Modeling Data]
</p>

@@

```
<p align = "center">
[Figure 2: Total Costs Boxplot]
</p>

@@

```{r, warning = FALSE, fig.height = 15, fig.width = 50}
#| echo: false
# Set the margin to include more space at the bottom
par(mar = c(5, 4, 4, 2) + 0.1)
par(cex.main = 5 )
cleaned_data$`Payment Typology 1` <- as.factor(cleaned_data$`Payment Typology 1`)
payments <- table(cleaned_data$`Payment Typology 1`)
pie(payments, main = "Payment Type Distribution", col = rainbow(length(payments)), labels = NA)
legend(1, 1, 1, legend = levels(cleaned_data$`Payment Typology 1`), fill = rainbow(length(payments)), cex = 3)

```
<p align = "center">
[Figure 3: Pie Chart of Payment Distribution]
</p>

@@

grid()
```
<p align = "center">
[Figure 4:Trends of Total Charges and Legnth of Stay ]
</p>

@@

```
<p align = "center">
[Figure 5: Age Group Distribution of Patients in New York]
</p>

@@

## Plot 6: Stacked Bar Chart by Gender and Race

@@

plot7 <- ggplot(cleaned_data, aes(x = Gender, fill = Race)) +
  geom_bar() +
  labs(
    title = "Stacked Bar Chart by Gender and Race",
    x = "Gender",
    y = "Count"
  ) +
  scale_fill_manual(
    values = c("White" = "lightblue", "Black" = "lightcoral", "Other" = "lightgreen"),
    breaks = c("White", "Black", "Other"),
    labels = c("White", "Black", "Other"),
    name = "Race"
  ) +
  theme_minimal()

plot7 + ggtitle("Stacked Bar Chart by Gender and Race")

```
<p align = "center">
[Figure 6: Stacked Bar Chart by Race and Gender]
</p>

@@

## Plot 7: Boxplots on the Total Costs on the Risk of Mortality

@@

```
<p align = "center">
[Figure 7: Boxplots on Total Cost and Mortality]
</p>

In Figure 7, our objective was to assess the central tendency of total hospital costs across different levels of Risk of Mortality. Employing separate box plots for each severity category allowed us to discern distinct patterns in the distribution of total costs associated with varying degrees of mortality risk. Notably, when focusing on Extreme cases of mortality, a clear trend emerged---median total costs were significantly higher, accompanied by a broader spread in the distribution. This observation implies that patients facing an Extreme risk of mortality not only incur higher median costs but also exhibit greater variability in the financial implications of their healthcare, suggesting a more diverse range of medical interventions and resource utilization.

This finding aligns with our initial expectation that higher levels of risk of mortality correspond to elevated costs of care. The discernible trend in the box plots reinforces the intuitive understanding that patients with a greater severity of illness often require more extensive and resource-intensive medical interventions, leading to increased healthcare costs. Beyond its immediate implications, this nuanced analysis offers valuable insights for healthcare planning and resource allocation, emphasizing the need for tailored strategies based on the specific needs associated with varying risk levels.

@@

## Plot 8: Violin Plot Comparing New York City and Western New York's Severity of Illness

@@

```
<p align = "center">
[Figure 8: NYC and Western NY Severity of Illness Comparsion ]
</p>

Figure 8 delves into a detailed breakdown of hospital patients, distinguishing between those originating from New York City (NYC) and Western New York (WNY). The focus is on the distribution of patients' APR severity codes, measured on a 1 to 4 scale (1 being mild, 2 being moderate, 3 being serious, and 4 being extreme). Notably, the violin plot reveals a higher frequency of APR severity code 2 assignments for patients from NYC compared to WNY. This divergence prompts exploration into potential variations in labeling practices between the two regions. However, beyond this distinction, the overall shapes of the violin distributions appear similar, suggesting a general alignment in severity profiles. This observation implies a degree of consistency in patient conditions based on APR severity codes, despite the regional differences in labeling practices. Understanding such variations is crucial for refining medical coding precision and ensuring uniformity in patient severity assessments across diverse geographical contexts.

@@

## Plot 9: Distribution of APR Severity of Illness for Medicaid Patients

@@

```
<p align = "center">
[Figure 9: APR Severity of Illness for Medicaid Patients]
</p>

In our examination of Figure 9, our focus shifted toward a specific subset of patients---those under Medicaid care. Medicaid, a health program designed to support individuals classified as economically disadvantaged, prompted our investigation into the severity of diagnoses within this demographic. Our objective was to discern whether patients under Medicaid receive comparable attention and care in contrast to their counterparts outside of the program.

@@

## Plot 10: Facet Density Plots of Total Cost per Age-Group Buckets

gg + scale_y_continuous(labels = scales::number_format(scale = 1e-3)) +
  labs(title = "Density Plot of Total Costs per Age-Group Bucket", x = "Total Costs", y = "Density (in thousands)") +
  theme_minimal()
```
<p align = "center">
[Figure 10: Density Plots of Total Costs per Age-Groups]
</p>

@@

In Figure 10, we analyzed age-specific total costs by charting density plots using a facet grid. For patients under 17, costs skewed right, suggesting lower expenses, potentially due to the overall good health of children. In contrast, as age increases, a broader and more dispersed peak in total costs emerges, indicating that older patients require more extensive and potentially costlier healthcare interventions, such as hospice services. This exploration sheds light on age-related patterns in healthcare costs, guiding strategies for resource allocation and care planning across different life stages.

@@

### References

New York State Department of Health. "Hospital Inpatient Discharges (SPARCS De-Identified)." data.ny.gov, 2013, https://health.data.ny.gov/Health/Hospital-Inpatient-Discharges-SPARCS-De-Identified/npsr-cm47.
